Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-05 paper/paper.qmd
Full discussion draft; fix inline hook

---
title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
date: last-modified
date-format: medium
prefer-html: true
bibliography: references.bib
fontsize: 11pt
format:
  manuscript-pdf:
    keep-tex: true
    keep-md: false
    linenumbers: true
    doublespacing: true
    lof: false
    header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
  html:
    embed-resources: true
runningtitle: Predicting CDI Severity from OTUs
runningauthor: Sovacool
papertype: Preprint
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    orcid: 0000-0002-6935-4275
    email: pschloss@umich.edu
    corresponding: true
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
abstract: |
  _Clostridioides difficile_ infection (CDI) can lead to 
  adverse outcomes including ICU admission, colectomy, and death.
  The composition of the gut microbiome plays an important role in determining
  colonization resistance and clearance upon exposure to _C. difficile_.
  We investigated whether machine learning (ML) models trained on gut microbiota
  isolated from stool samples on the day of CDI diagnosis could predict which
  CDI cases led to severe outcomes.
  We collected 1,277 stool samples from CDI patients and clustered 16S rRNA gene
  amplicon sequences into _de novo_ Operational Taxonomic Units (OTUs).
  We then trained ML models to predict CDI severity on OTU relative abundances
  according to four different severity definitions: 
  the IDSA severity score on the day of diagnosis,
  all-cause adverse outcomes within 30 days,
  adverse outcomes confirmed as attributable to CDI via chart review, 
  and a pragmatic definition that uses the attributable definition when available
  and falls back to the all-cause definition otherwise.
  The models predicting pragmatic severity performed best, 
  suggesting that while chart review is valuable to verify the cause of
  complications,
  including as many samples as possible is indispensable for training performant
  models on imbalanced datasets.
  Permutation importance identified _Enterococcus_ as the most important OTU for
  model performance. 
  Finally, we evaluated the potential clinical value of the OTU-based models
  and found similar performance compared to prior models based on Electronic
  Health Records.
  The modest performance of the OTU-based models represents a step toward the
  goal of deploying models to inform clinical decisions and ultimately improve
  CDI outcomes.
#importance: TODO
keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
execute:
  eval: true
  echo: false
  cache: true
  message: false
  warning: false
---
```{r setup}
#| include: false
schtools::set_knitr_opts()
knitr::opts_chunk$set(dpi = 600)
```
```{r load_data}
#| output: false
library(assertthat)
library(glue)
library(here)
library(kableExtra)
library(knitr)
library(tidyverse)
load(here("results", "stats.RData"))
table_font_size <- 10
assert_that(n_cases_first == 1277) # hard-coded in abstract
```

@@

We performed permutation feature importance to determine which OTUs contributed
the most to model performance.
An OTU was considered important if performance decreased when it was permuted in
at least 75% of the train/test splits, with greater differences in AUROC meaning
greater importance.
We plotted mean decrease in AUROC alongside log~10~-transformed mean
relative abundances for the top OTUs (@fig-features).
_Enterococcus_ was the most important OTU, being significantly important for 
all models except for attributable severity on the full dataset.
_Staphylococcus_ was important for the pragmatic and all-cause definitions on
the full datasets, but not for models trained on the intersection dataset.
_Lactobacillus_ was important only for the all-cause definition on the intersection
dataset.
All remaining OTUs had differences in AUROC < 0.02 and were only significantly
important in one or two of the models at most.
There is not always a clear pattern of increased or decreased relative abundance
for important OTUs in severe CDI cases, but all of the top 5 OTUs had an
increased mean relative abundance in severe cases relative to not severe cases.
TODO concluding sentence.

@@

We trained ML models based on gut microbial communities on the day of CDI
diagnosis to predict CDI severity according to four different severity definitions.
The purpose of using the full datasets is to train the best models possible
given the constraints, while using the intersection dataset allows for comparing
severity definitions.
We found that models predicting pragmatic severity with as much data as
available performed best, and that performance dropped substantially when 
reducing to the intersection datasets for all definitions likely due to the
particularly imbalanced nature of the all-cause and attributable definitions.
These results demonstrate the importance of using as many samples as possible when
data are sparse and the outcome is low prevalence, as well as the need to
incorporate physician's expertise when possible.

@@

The full pragmatic severity model performed just as well as a prior
EHR-based model trained on the day of diagnosis, demonstrating the potential
utility of OTU-based models.
However, it is not enough for models to perform well to justify deploying them
in a clinical setting; benefit over current practices must be shown.
We calculated the number needed to screen for our OTU-based models and compared
these to EHR-based and curated models trained two days after the diagnosis, and
found the OTU-based pragmatic severity model outperformed the curated models but
not the EHR models.
While the attributable definition had a worse NNS for the OTU-based models, it
did not perform worse than the prior curated model, and it may be the most
clinically relevant as physician chart review increases confidence that
positively-labelled severe outcomes are due to the CDI rather than other causes.

Although no known treatment options are proven to reduce the risk of severe CDI
outcomes, fidaxomicin is promising due to its improved time to resolution and
reduced recurrence.
Fidaxomicin is also attractive as a preferential antibiotic option as
vancomycin-resistant enterococcus is on the rise and enterococci are known to
worsen CDI [@poduval_clostridium_2000;@smith_enterococci_2022].
We extended our analysis of clinical value to incorporate the number needed to
treat for fidaxomicin alongside the predictive models in order to calculate the
number needed to benefit.
The NNB contextualizes model performance within clinical reality.

A more robust analysis of clinical value would further consider the cost of
treatment options versus the savings of averting severe outcomes, as economic
disparities are a major barrier to treatment in the US.
Cost-benefit analyses have found that fidaxomicin is more cost-effective than
vancomycin largely due to the reduced risk of recurrence [@johnson_clinical_2021;@jiang_budget_2022].

As patient age is already the most significant risk factor, the gut microbiome
may only a minor part in influencing CDI severity.
However, deploying models incorporating both microbial factors and patient
factors from the EHR could be worth the time and costs if
performance improvements were shown translate to clinically significant improvements in
patient outcomes.
in practice, EHR-based models are easier and less costly to deploy than an OTU-based model would be -- can be embedded for automated scores.
further work needed to characterize the costs and benefits.

Amplicon sequencing is not typically performed for CDI patients, but if there is
clinical value to be gained by implementing OTU-based models, routinely
sequencing and profiling the microbial communities of CDI patients could be
justified.

Models predicting the pragmatic definition yielded the best NNS.
Future work is needed to determine whether incorporating both microbial features
and EHRs would result in a clinically meaningful improvement in performance over
deploying EHR models alone.

Two goals: investigate whether we can predict CDI severity based on OTU data to
inform how the gut microbiome may modulate severity (ML-based science: good
performance implies something about underlying biology), and determine whether
there is potential clinical value in OTU-based models.
new dataset.

<!--
we do not know which antibiotics were prescribed to treat these CDI cases, nor
which antibiotics patients may have taken prior to the CDI.
differences in microbiota between patients may be due to different abx...
different antibiotics have been shown to create different forms of dysbiotic
microbiota [@berkell_microbiota-based_2021] and
differential cdi clearance [@tomkovich_initial_2020].

>>>>

---
title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
date: last-modified
date-format: medium
prefer-html: true
bibliography: references.bib
fontsize: 11pt
format:
  manuscript-pdf:
    keep-tex: true
    keep-md: false
    linenumbers: true
    doublespacing: true
    lof: false
    header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
  html:
    embed-resources: true
runningtitle: Predicting CDI Severity from OTUs
runningauthor: Sovacool
papertype: Preprint
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    orcid: 0000-0002-6935-4275
    email: pschloss@umich.edu
    corresponding: true
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
abstract: |
  _Clostridioides difficile_ infection (CDI) can lead to 
  adverse outcomes including ICU admission, colectomy, and death.
  The composition of the gut microbiome plays an important role in determining
  colonization resistance and clearance upon exposure to _C. difficile_.
  We investigated whether machine learning (ML) models trained on gut microbiota
  isolated from stool samples on the day of CDI diagnosis could predict which
  CDI cases led to severe outcomes.
  We collected 1,277 stool samples from CDI patients and clustered 16S rRNA gene
  amplicon sequences into _de novo_ Operational Taxonomic Units (OTUs).
  We then trained ML models to predict CDI severity on OTU relative abundances
  according to four different severity definitions: 
  the IDSA severity score on the day of diagnosis,
  all-cause adverse outcomes within 30 days,
  adverse outcomes confirmed as attributable to CDI via chart review, 
  and a pragmatic definition that uses the attributable definition when available
  and falls back to the all-cause definition otherwise.
  The models predicting pragmatic severity performed best, 
  suggesting that while chart review is valuable to verify the cause of
  complications,
  including as many samples as possible is indispensable for training performant
  models on imbalanced datasets.
  Permutation importance identified _Enterococcus_ as the most important OTU for
  model performance. 
  Finally, we evaluated the potential clinical value of the OTU-based models
  and found similar performance compared to prior models based on Electronic
  Health Records.
  The modest performance of the OTU-based models represents a step toward the
  goal of deploying models to inform clinical decisions and ultimately improve
  CDI outcomes.
#importance: TODO
keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
execute:
  eval: true
  echo: false
  cache: true
  message: false
  warning: false
---
```{r setup}
#| include: false
schtools::set_knitr_opts()
```
```{r load_data}
#| output: false
library(assertthat)
library(glue)
library(here)
library(kableExtra)
library(knitr)
library(tidyverse)
load(here("results", "stats.RData"))
table_font_size <- 10
assert_that(n_cases_first == 1277) # hard-coded in abstract
```

@@

We performed permutation feature importance to determine which OTUs contributed
the most to model performance.
An OTU was considered important if performance decreased when it was permuted in
at least 75% of the train/test splits, with greater differences in AUROC meaning
greater importance.
We plotted mean decrease in AUROC alongside log~10~-transformed mean
relative abundances for the top OTUs (@fig-features).
_Enterococcus_ was the most important OTU, being significantly important for 
all models except for attributable severity on the full dataset.
_Staphylococcus_ was important for the pragmatic and all-cause definitions on
the full datasets, but not for models trained on the intersection dataset.
_Lactobacillus_ was important only for the all-cause definition on the intersection
dataset.
All remaining OTUs had differences in AUROC < 0.02 and were only significantly
important in one or two of the models at most.
There is not always a clear pattern of increased or decreased relative abundance
for important OTUs in severe CDI cases, but all of the top 5 OTUs had an
increased mean relative abundance in severe cases relative to not severe cases.

@@

We trained ML models based on gut microbial communities on the day of CDI
diagnosis to predict CDI severity according to four different severity definitions.
The purpose of using the full datasets is to train the best models possible
given the constraints, while using the intersection dataset allows for comparing
severity definitions.
We found that models predicting pragmatic severity with as much data as
available performed best, and that performance dropped substantially when 
reducing to the intersection datasets for all definitions, likely due to the
particularly imbalanced nature of the all-cause and attributable definitions.
These results demonstrate the importance of using as many samples as possible when
data are sparse and the outcome is low prevalence, as well as the need to
incorporate physician's expertise when possible.

@@

The full pragmatic severity model performed just as well as a prior
EHR-based model trained on the day of diagnosis, demonstrating the potential
utility of OTU-based models.
In terms of the number needed to screen, the OTU-based pragmatic severity model 
outperformed a prior model of manually curated clinical variables, but not a
model trained on EHR data extracted two days after the diagnosis.
While the attributable definition had a worse NNS for the OTU-based models, it
did not perform worse than the prior curated model, and it may be the most
clinically relevant as physician chart review increases confidence that
positively-labelled severe outcomes are due to the CDI rather than other causes.

However, it is not enough for models to perform well to justify deploying them
in a clinical setting; benefit over current practices must be shown [@wiens_no_2019].
Although no known treatment options are proven to reduce the risk of severe CDI
outcomes, fidaxomicin is promising due to its improved time to resolution and
reduced recurrence.
Fidaxomicin is also attractive as a preferential antibiotic option as
vancomycin-resistant enterococcus is on the rise and enterococci are known to
worsen CDI [@poduval_clostridium_2000;@smith_enterococci_2022].
We extended our analysis of clinical value to incorporate the number needed to
treat for fidaxomicin alongside the predictive models in order to calculate the
number needed to benefit.
The NNB contextualizes model performance within clinical reality, as it combines
both model performance and treatment performance [@liu_number_2019].
A more robust analysis of clinical value would further consider the cost of
treatment options versus the savings of averting severe outcomes, as economic
disparities are a major barrier to treatment in the US.
Cost-benefit analyses have found that fidaxomicin is more cost-effective than
vancomycin, largely due to the reduced risk of recurrence [@johnson_clinical_2021;@jiang_budget_2022].
While our analysis of clinical value is only a proof-of-concept, if evidence
emerges that new or existing treatments significantly reduce the risk of severe
CDI, our results can be incorporated into future considerations of whether to
build severity prediction models and what features should be incorporated.
In practice, EHR-based models are less costly to deploy than OTU-based models
and do not require additional clinical sample collection.
Amplicon sequencing is not typically performed for CDI patients, however,
routinely sequencing and profiling the microbial communities of CDI patients
could be justified if models that incorporate microbial features were shown to
improve patient outcomes.

In all, we found that our models to predict severity from features of the gut
microbiome performed moderately well.
Our approach enabled us to identify bacteria that contributed to model
performance and evaluate how well the state of the gut microbiome can predict
several different definitions of CDI severity.
Further work is needed to determine whether the performance of OTU-based models
is sufficient to justify their deployment in clinical settings, especially as
compared to EHR-based models.
If and when new evidence emerges of improved treatments to prevent severe CDI
outcomes, deploying performant and robust models for clinicians to tailor
treatment options promises to reduce morbidity and mortality of CDI.
<!--
we do not know which antibiotics were prescribed to treat these CDI cases, nor
which antibiotics patients may have taken prior to the CDI.
differences in microbiota between patients may be due to different abx...
different antibiotics have been shown to create different forms of dysbiotic
microbiota [@berkell_microbiota-based_2021] and
differential cdi clearance [@tomkovich_initial_2020].
